Linking watershed nutrient loading to estuary water quality with semiparametric models

### 2023-05-31 Schramm-2023-06-PeerJ/Schramm-2023-06-PeerJ.tex
trends plot plus lots of edits

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-basic,referee,lineno,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\raggedbottom

@@

@@

\title[Watershed nutrient loading and water quality in a subtropical
estuary]{Assessing linkages between watershed nutrient loading and water
quality in a subtropical estuary with semiparametric models.}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Michael} \sur{Schramm} }\email{\href{mailto:michael.schramm@ag.tamu.edu}{\nolinkurl{michael.schramm@ag.tamu.edu}}}

  \affil*[1]{\orgdiv{Texas Water Resources Institute}, \orgname{Texas
A\&M AgriLife Research}, \orgaddress{\city{College
Station}, \country{United
States}, \postcode{77840}, \state{TX}, \street{1001 Holleman Dr.~E.}}}

\abstract{Lavaca Bay is a secondary embayment on the Texas coast
displaying early signals of water quality degradation. This study
applied a semiparametric approach to assess both watershed nutrient
loads and responses in estuary water quality to nutrient loading and
streamflow. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated there was no evidence of long-term changes in
dissolved oxygen or chlorophyll-\emph{a} within Lavaca Bay. However,
site specific long-term increases in both organic and inorganic nitrogen
are concerning. Further analysis found freshwater inflow was a strong
driver of nutrient and chlorophyll-\emph{a} concentrations within Lavaca
Bay but there was no evidence that changes in watershed nutrient loading
explained variation in dissolved oxygen or chlorophyll-\emph{a}
concentrations. In addition to providing a baseline assessment of
watershed nutrient loading and water quality responses in the Lavaca Bay
watershed, this study provides methodological support for the use of
semiparametric methods in load regression models and estuary
assessments.}

\keywords{key, dictionary, word}

\maketitle

\emph{Remaining Items}

\begin{itemize}
\tightlist
\item
  keywords
\item
  discussion
\item
  formalize language (reduce first person active?)
\item
  final language review
\item
  springer doesn't want figures in subdirectory, need to knit to same
  directory as tex file.
\end{itemize}

\hypertarget{sec1}{%
\section{Introduction}\label{sec1}}

Similar to many coastal areas globally, the coastal watersheds along the
Texas Gulf coast are facing pressures from increasing population,
increases in point source and non-point source pollution and alterations
to freshwater flows that degrade water quality in downstream estuaries
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Despite these escalating pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or low risk for
eutrophic conditions \citep{bricker_effects_2008}. However, a suite of
recent studies indicates that estuary water quality dynamics in both
agriculturally dominated and urban watersheds within Texas are
expressing conditions that are increasingly conducive to algal blooms
and eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.
With identification of several localized areas of estuary water quality
concern along the Texas coast \citep{bugica_water_2020}, localized
studies are being prioritized to better inform management actions.

This project aims to provide an assessment of watershed nutrient loading
and resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay
is a secondary bay in the larger Matagorda Bay system located roughly
halfway between Houston, Texas and Corpus Christi, Texas. Lavaca Bay
faces substantial challenges associated with legacy contamination but
general water quality parameters such as dissolved oxygen (DO),
nutrients, and biological parameters have been well within state water
quality standards. Despite largely meeting statewide water quality
standards, there have been concerning recently identified declines in
abundance, biomass, and diversity of benthic fauna in Lavaca Bay
\citep{beserespollackLongtermTrendsResponse2011}. These declines are
partially attributed to reductions in freshwater inflow and changes in
estuary salinity and are indicative of an already stressed system
\citep{beserespollackLongtermTrendsResponse2011, palmerImpactsDroughtsLow2015, montagnaAssessmentRelationshipFreshwater2020}.
More recently, significant linear increases in total phosphorus (TP),
orthophosphate, total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a}
concentrations were identified at monitoring sites within Lavaca Bay
\citep{bugica_water_2020}. Although long-term changes in DO
concentrations have not been identified, the trends in nutrient
concentrations are concerning due to the role of nitrogen as a limiting
factor for primary production in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\citep{jainTechnicalSupportDocument2021, schrammLavacaRiverWatershed2018, bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to identify and
replicate effective management actions across the state
\citep{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\citep{wazniakLinkingWaterQuality2007, lloydMethodsDetectingChange2014}.
The development and application of flexible statistical methods such as
Weighted Regressions on Time, Discharge and Season
\citep[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \citep[GAMs,][]{woodFastStableRestricted2011} have
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in-stream concentrations or loads
(daily, monthly, or annually) and flow-normalized estimates of
concentrations and loads using locally weighted regresssion for unique
combinations of time, discharge, and season. WRTDS has been widely used
to assess and identify trends in riverine nutrients
\citep{oelsner_recent_2019, stackpooleLongTermMississippi2021},
chlorides \citep{stetsIncreasingChlorideRivers2018}, and other
pollutants of concern \citep{shodaWaterqualityTrendsRivers2019}. WRTDS
has also been succesfully adapted to assess trends in estuarine water
quality concentrations \citep{beckFourDecadesWater2018}.

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\citep{woodFastStableRestricted2011}. Although the underlying parameter
estimation procedure of GAMs is substantially different than WRTDS, both
the functional form and results have been demonstrated to be similar
when assessing nutrient concentration trends
\citep{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs have included include river and catchment nutrient
concentration and load estimation
\citep{wangLoadEstimationUncertainties2011, kroonRiverLoadsSuspended2012, kuhnertQuantifyingTotalSuspended2012, robson_prediction_2015-1, hagemannEstimatingNutrientOrganic2016, mcdowell_implications_2021, biagi_novel_2022},
assessment of temporal trends of nutrients
\citep{beckNumericalQualitativeContrasts2017, murphyGeneralizedAdditiveModel2019},
phytoplankton \citep{bergbuschUnexpectedShiftPhytoplankton2021}, and
cyanobacteria \citep{hayesEffectsLakeWarming2020}. Recently GAMs have
also been used to link water quality responses in receiving water bodies
to changes in nonpoint source nutrient inputs
\citep{murphyNutrientImprovementsChesapeake2022}.
\citet{beckNumericalQualitativeContrasts2017} provides a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications.

To provide actionable information for resource managers in Lavaca Bay,
water quality conditions must be evaluated relative to changes in
natural environmental drivers to better understand and manage potential
human impacts. This study utilizes GAMs to develop estimates of
delivered and flow-normalized nutrient loads and assess changes in loads
delivered to Lavaca Bay. GAMs were chosen over other regression-based
approached for use in this study due to; (1) the ability to easily
explore and incorporate different model terms; (2) the incorporation of
non-liner smooth functions that do not require explicit a priori
knowledge of the expected shape; and (3) inclusion of a link function
that related the expected value of the response to linear predictors
thus avoiding unneeded data transformations and bias corrections. The
exploratory study also assesses the response of water quality parameters
in Lavaca Bay over time and in response to freshwater inflow controlled
for seasonality and to watershed nutrient loads that are controlled for
environmentally driven variation.

\hypertarget{sec2}{%
\section{Methods}\label{sec2}}

\hypertarget{location-and-data}{%
\subsection{Location and Data}\label{location-and-data}}

Lavaca Bay is 190 km\textsuperscript{2} with the majority of freshwater
inflow provided by the Lavaca and Navidad River systems
(Figure~\ref{fig:fig1}). The Garcitas-Arenosa, Placedo Creek, and Cox
Bay watersheds provide additional freshwater inflows. The entire
watershed land area is 8,149 km\textsuperscript{2} and primarily rural.
Watershed land cover and land use is 50\% grazed pasture and rangeland,
20\% cultivated cropland (primarily rows crops such as corn, cotton, and
sorghum), and 5\% suburban/urban. Pasture and rangeland is concentrated
in the Lavaca River watershed, while cultivated crops are generally
located along the eastern tributaries of the Navidad river. The Lavaca
and Navidad River watersheds are a combined 5,966 km\textsuperscript{2},
or approximately 73\% of the entire Lavaca Bay watershed area. Discharge
from the Navidad River is regulated by Lake Texana which has been in
operation since 1980. Lake Texana provides 0.210 km\textsuperscript{3}
of water storage and discharges into the tidal section of the Navidad
River which ultimately joins the tidal section of the Lavaca River 15 km
upstream of the confluence with the Lavaca Bay.

@@

Daily discharges for the Lavaca River (USGS-08164000,
Figure~\ref{fig:fig1}) were obtained from the United States Geologic
Survey (USGS) National Water Information System using the
\emph{dataRetrieval} R package
\citep{deciccoDataRetrievalPackagesDiscovering2022}. Gaged daily
discharges from the outlet of Lake Texana on the Navidad River
(USGS-0816425) were provided by the Texas Water Development Board (TWDB)
(April 21, 2022 email from R. Neupane, TWDB).

Water quality sample data for the two freshwater and three estuary
locations were obtained from the Texas Commission on Environmental
Quality (TCEQ) Surface Water Quality Monitoring Information System. Data
submitted through the system are required to be collected under Quality
Assurance Project Plans and lab method procedures outlined by the TCEQ's
procedures manual to ensure consistent collection and laboratory methods
are applied between samples collected by different entities and under
different projects. All sites had varying lengths of and availability of
data. For freshwater locations, TP from January 2000 through December
2020 and nitrate-nitrogen (NO\textsubscript{3}) data from January 2005
through December 2020 were downloaded (Table~\ref{tab:table1}). Less
than 5-years of total nitrogen and TKN concentration data were available
at the freshwater sites and deemed insufficient to develop load
estimation models
\citep{horowitzEvaluationSedimentRating2003, snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\emph{\textsubscript{x}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}).

@@

\hypertarget{estimating-watershed-based-nutrient-loads}{%
\subsection{Estimating Watershed Based Nutrient
Loads}\label{estimating-watershed-based-nutrient-loads}}

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164525) were developed using GAMs relating nutrient
concentration to river discharge, season, and time. Separate models were
fit at each station for each parameter and used to predict nutrient
concentrations for each day in the study period. GAMs were fit using the
\emph{mgcv} package in R which makes available multiple types of smooth
functions with automatic smoothness selection
\citep{woodFastStableRestricted2011}. The general form of the model
related NO\textsubscript{3} or TP concentration to a long term tend,
season, streamflow, and two different antecedent discharge terms, shown
in Eq. \ref{eq1}:

@@

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period
\citep{wangLoadEstimationUncertainties2011, kuhnertQuantifyingTotalSuspended2012, zhang_improving_2017},
using Eq. \ref{eq2}:

\begin{align}
ma(\delta) &= d{\kappa_{i-1}}+(1-\delta)\hat{q}_{i-1}, \nonumber \\
\kappa_{i} &=\sum_{m=1}^{i}\hat{Q}_m
\label{eq2}
\end{align}

where \(\delta\) is the discount factor (here, set equal to 0.95),
\(\kappa_i\) is the cumulative flow (\emph{Q}) up to the \emph{i}th day.

Flow anomaly (\emph{fa}) is a unitless term that represents how wet or
dry the current time period is from a previous time period
\citep{vecchiaTrendsConcentrationsUse2009, zhang_improving_2017}.
Long-term flow anomaly (\emph{ltfa}) is the streamflow over the previous
year relative to the entire period \citep{zhang_improving_2017} and
calculated using Eq. \ref{eq3}:

\begin{equation}
ltfa(t) = \bar{x}_{1\,year}(t) - \bar{x}_{entire\,period}
\label{eq3}
\end{equation}

and the short-term flow anomaly (\emph{stfa}) calculated as the current
day flow compared to the preceding 1-month streamflow using \ref{eq4}:

\begin{equation}
stfa(t) = x_{current\,day}(t) - \bar{x}_{1\,month}(t)
\label{eq4}
\end{equation}

where \(x\) are the averages of log-transformed streamflow over the
antecedent period (\emph{1-year}, \emph{1-month}, etc.) for time
\emph{t}. We used \emph{ltfa} in NO\textsubscript{3} models and
\emph{stfa} in TP models based on previous work demonstrating major
improvements in NO\textsubscript{x} regression models that incorporated
\emph{ltfa} and moderate improvements in TP regression models that
incorporated \emph{stfa} \citep{zhang_improving_2017}. Moving averages
and flow anomalies were calculated with the \emph{adc} R package
\citep{schrammAdcCalculateAntecedant2023}.

@@

Left-censored data were not uncommon in this dataset. Several methods
are available to account for censored data. We transformed left-censored
nutrient concentrations to one-half the detection limit. Although this
simple approach can introduce bias
\citep{hornungEstimationAverageConcentration1990}, we considered it
acceptable because high concentrations and loadings are associated with
high-flow events and low-flow/low-concentration events will account for
a small proportion of total loadings \citep{mcdowell_implications_2021}.

>>>>

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

% ZNK -- Adding headers for pandoc

\setlength{\emergencystretch}{3em}
\usepackage[unicode=true]{hyperref}

@@

\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Linking watershed nutrient loading to estuary water quality with
semiparametric models}

\author[1]{Michael Schramm}

\corrauthor[1]{Michael
Schramm}{\href{mailto:michael.schramm@ag.tamu.edu}{\nolinkurl{michael.schramm@ag.tamu.edu}}}

\affil[1]{Texas Water Resources Institute, Texas A\&M AgriLife Research,
College Station, Texas, United States}

%
% \author[1]{First Author}
% \author[2]{Second Author}
% \affil[1]{Address of first author}
% \affil[2]{Address of second author}
% \corrauthor[1]{First Author}{f.author@email.com}

% 
\usepackage[style=apa,]{biblatex}
\addbibresource{bibliography.bib}

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
semiparametric approaches to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degredation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by non-natural
variation. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are causes of
concern as potential causes of future eutrophication. Further analysis
found freshwater inflow was a strong driver of nutrient and
chlorophyll-\emph{a} concentrations within Lavaca Bay but there was no
evidence that changes in watershed nutrient loading explained additional
variation in dissolved oxygen or chlorophyll-\emph{a} concentrations. In
addition to providing a baseline assessment of watershed nutrient
loading and water quality responses in the Lavaca Bay watershed, this
study provides methodological support for the use of semiparametric
methods in load regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

\flushbottom
\maketitle
\thispagestyle{empty}

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Similar to many coastal areas globally, the coastal watersheds along the
Texas Gulf coast are facing pressures from increasing population,
increases in point source and non-point source pollution and alterations
to freshwater flows that degrade water quality in downstream estuaries
\autocite{bricker_effects_2008,kennicuttWaterQualityGulf2017,bugica_water_2020}.
Despite these escalating pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or low risk for
eutrophic conditions \autocite{bricker_effects_2008}. However, a suite
of recent studies indicates that estuary water quality dynamics in both
agriculturally dominated and urban watersheds within Texas are
expressing conditions that are increasingly conducive to algal blooms
and eutrophication
\autocite{wetzWaterQualityDynamics2016,wetz_exceptionally_2017,bugica_water_2020,chinPhytoplanktonBiomassCommunity2022}.
With identification of several localized areas of estuary water quality
concern along the Texas coast \autocite{bugica_water_2020}, localized
studies are being prioritized to better inform management actions.

This goal of this project is to assess watershed nutrient loading and
the resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay
is a secondary bay in the larger Matagorda Bay system located roughly
halfway between Houston, Texas and Corpus Christi, Texas. Lavaca Bay
faces substantial challenges associated with legacy contamination but
general water quality parameters such as dissolved oxygen (DO),
nutrients, and biological parameters have been well within state water
quality standards. Despite largely meeting statewide water quality
standards, there have been concerning declines in abundance, biomass,
and diversity of benthic fauna in Lavaca Bay
\autocite{beserespollackLongtermTrendsResponse2011}. These declines are
partially attributed to reductions in freshwater inflow and changes in
estuary salinity and are indicative of an already stressed system
\autocite{beserespollackLongtermTrendsResponse2011,palmerImpactsDroughtsLow2015,montagnaAssessmentRelationshipFreshwater2020}.
More recently, significant linear increases in total phosphorus (TP),
orthophosphate, total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a}
concentrations were identified at monitoring sites within Lavaca Bay
\autocite{bugica_water_2020}. Although long-term changes in DO
concentrations have not been identified, the trends in nutrient
concentrations are concerning due to the role of nitrogen as a limiting
factor for primary production in many Texas estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\autocite{jainTechnicalSupportDocument2021,schrammLavacaRiverWatershed2018,bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to identify and
replicate effective management actions across the state
\autocite{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\autocite{wazniakLinkingWaterQuality2007,lloydMethodsDetectingChange2014}.
The development and application of flexible statistical methods such as
Weighted Regressions on Time, Discharge and Season
\autocite[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \autocite[GAMs,][]{woodFastStableRestricted2011} have
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in-stream concentrations or loads
(daily, monthly, or annually) and flow-normalized estimates of
concentrations and loads using locally weighted regresssion for unique
combinations of time, discharge, and season. WRTDS has been widely used
to assess and identify trends in riverine nutrients
\autocite{oelsner_recent_2019,stackpooleLongTermMississippi2021},
chlorides \autocite{stetsIncreasingChlorideRivers2018}, and other
pollutants of concern \autocite{shodaWaterqualityTrendsRivers2019}.
WRTDS has also been succesfully adapted to assess trends in estuarine
water quality concentrations \autocite{beckFourDecadesWater2018}.

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\autocite{woodFastStableRestricted2011}. Although the underlying
parameter estimation procedure of GAMs are substantially different than
WRTDS, both the functional form and results have been demonstrated to be
similar when assessing nutrient concentration trends
\autocite{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs have included river and catchment nutrient
concentration and load estimation
\autocite{wangLoadEstimationUncertainties2011,kroonRiverLoadsSuspended2012,kuhnertQuantifyingTotalSuspended2012,robson_prediction_2015-1,hagemannEstimatingNutrientOrganic2016,mcdowell_implications_2021,biagi_novel_2022},
and assessment of temporal trends of nutrients
\autocite{beckNumericalQualitativeContrasts2017,murphyGeneralizedAdditiveModel2019},
phytoplankton \autocite{bergbuschUnexpectedShiftPhytoplankton2021}, and
cyanobacteria \autocite{hayesEffectsLakeWarming2020}. Recently GAMs have
also been used to link water quality responses in receiving water bodies
to changes in nonpoint source nutrient inputs
\autocite{murphyNutrientImprovementsChesapeake2022}. For a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications reders are referred to
\textcite{beckNumericalQualitativeContrasts2017}.

To provide actionable information for resource managers in Lavaca Bay,
water quality conditions must be evaluated relative to changes in
natural environmental drivers to better understand and manage potential
human impacts. This study utilizes GAMs to develop estimates of
delivered and flow-normalized nutrient loads and assess estuary water
quality responses to changes in loads delivered to Lavaca Bay. GAMs were
chosen over other regression-based approached for use in this study due
to; (1) the ability to easily explore and incorporate different model
terms; (2) the incorporation of non-linear smooth functions that do not
require explicit a priori knowledge of the expected shape; and (3)
inclusion of a link function that related the expected value of the
response to linear predictors thus avoiding unneeded data
transformations and bias corrections. The exploratory study also
assesses the response of water quality parameters in Lavaca Bay over
time and in response to freshwater inflow controlled for seasonality and
to watershed nutrient loads that are controlled for environmentally
driven variation.

\hypertarget{methods-and-materials}{%
\section*{Methods and Materials}\label{methods-and-materials}}
\addcontentsline{toc}{section}{Methods and Materials}

\hypertarget{location-and-data}{%
\subsection*{Location and Data}\label{location-and-data}}
\addcontentsline{toc}{subsection}{Location and Data}

Lavaca Bay is a 190 km\textsuperscript{2} estuary with the majority of
freshwater inflow provided by the Lavaca and Navidad River systems
(Fig.~\ref{fig:fig1}). The Garcitas-Arenosa, Placedo Creek, and Cox Bay
subwatersheds provide additional freshwater inflows. The entire
watershed area is 8,149 km\textsuperscript{2} and primarily rural.
Watershed land cover and land use is 50\% grazed pasture and rangeland,
20\% cultivated cropland (primarily rows crops such as corn, cotton, and
sorghum), and 5\% suburban/urban. Pasture and rangeland is concentrated
in the Lavaca River watershed, while cultivated crops are generally
located along the eastern tributaries of the Navidad river. The Lavaca
and Navidad River watersheds are a combined 5,966 km\textsuperscript{2},
or approximately 73\% of the entire Lavaca Bay watershed area. Discharge
from the Navidad River is regulated by Lake Texana which has been in
operation since 1980. Lake Texana provides 0.210 km\textsuperscript{3}
of water storage and discharges into the tidal section of the Navidad
River which ultimately joins the tidal section of the Lavaca River 15 km
upstream of the confluence with the Lavaca Bay.

@@

Daily discharges for the Lavaca River (USGS-08164000,
Fig.~\ref{fig:fig1}) were obtained from the United States Geologic
Survey (USGS) National Water Information System using the
\emph{dataRetrieval} R package
\autocite{deciccoDataRetrievalPackagesDiscovering2022}. Gaged daily
discharges from the outlet of Lake Texana on the Navidad River
(USGS-0816425) were provided by the Texas Water Development Board (TWDB)
(April 21, 2022 email from R. Neupane, TWDB).

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual to ensure consistent collection and laboratory methods
are applied between samples collected by different entities and under
different projects. All sites had varying lengths of and availability of
data. For freshwater locations, TP from January 2000 through December
2020 and nitrate-nitrogen (NO\textsubscript{3}) data from January 2005
through December 2020 were downloaded (Table~\ref{tab:table1}). Less
than 5-years of total nitrogen and TKN concentration data were available
at the freshwater sites and deemed insufficient to develop load
estimation models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}).

@@

\hypertarget{estimating-watershed-based-nutrient-loads}{%
\subsection*{Estimating Watershed Based Nutrient
Loads}\label{estimating-watershed-based-nutrient-loads}}
\addcontentsline{toc}{subsection}{Estimating Watershed Based Nutrient
Loads}

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164525) were developed using GAMs relating nutrient
concentration to river discharge, season, and time. Separate models were
fit at each station for each parameter and used to predict nutrient
concentrations for each day in the study period. GAMs were fit using the
\emph{mgcv} package in R which makes available multiple types of smooth
functions with automatic smoothness selection
\autocite{woodFastStableRestricted2011}. The general form of the model
related NO\textsubscript{3} or TP concentration to a long term tend,
season, streamflow, and two different antecedent discharge terms, shown
in Eq. \ref{eq1}:

@@

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period
\autocite{wangLoadEstimationUncertainties2011,kuhnertQuantifyingTotalSuspended2012,zhang_improving_2017},
using Eq. \ref{eq2}:

\begin{equation}
ma(t) = \frac{\sum^{J}_{j=1}{d^{t-t_j}Q_{t_{j}}}}{\sum^{J}_{j=1}d^{t-t_j}}
\label{eq2}
\end{equation}

where \emph{t} is the day of estimation, \emph{d} is the discount factor
(set to 0.95), \emph{J} is the total number of daily discharge
observations, \(t-t_j\) is the lag time between the current day and some
historical day of observation (\(t_j\)). Here, as \emph{d} approaches
zero, the time series becomes the daily observed values and as \emph{d}
gets closer to one, the time series becomes the mean flow.

Flow anomaly (\emph{fa}) is a unitless term that represents how wet or
dry the current time period is from a previous time period
\autocite{vecchiaTrendsConcentrationsUse2009,zhang_improving_2017}.
Long-term flow anomaly (\emph{ltfa}) is the streamflow over the previous
year relative to the entire period
\autocite[Eq.~\ref{eq3},][]{zhang_improving_2017} and and the short-term
flow anomaly (\emph{stfa}) calculated as the current day flow compared
to the preceding 1-month streamflow
\autocite[Eq.~\ref{eq4},][]{zhang_improving_2017}:

\begin{align}
ltfa(t) &= \bar{x}_{1\,year}(t) - \bar{x}_{entire\,period}, \label{eq3}\\
stfa(t) &= x_{current\,day}(t) - \bar{x}_{1\,month}(t) \label{eq4}
\end{align}

where \(x\) are the averages of log-transformed streamflow over the
antecedent period (\emph{1-year}, \emph{1-month}, etc.) for time
\emph{t}. We used \emph{ltfa} in NO\textsubscript{3} models and
\emph{stfa} in TP models based on previous work demonstrating major
improvements in NO\textsubscript{\emph{X}} regression models that
incorporated \emph{ltfa} and moderate improvements in TP regression
models that incorporated \emph{stfa} \autocite{zhang_improving_2017}.
Moving averages and flow anomalies were calculated with the \emph{adc} R
package \autocite{schrammAdcCalculateAntecedant2023}.

@@

Left-censored data were not uncommon in this dataset. Several methods
are available to account for censored data. We transformed left-censored
nutrient concentrations to one-half the detection limit. Although this
simple approach can introduce bias
\autocite{hornungEstimationAverageConcentration1990}, we considered it
acceptable because high concentrations and loadings are associated with
high-flow events and low-flow/low-concentration events will account for
a small proportion of total loadings
\autocite{mcdowell_implications_2021}.
